Denotational Semantics of Gradual Typing using Synthetic Guarded Domain Theory

### 2024-01-27 paper-new/categorical-model.tex
squeeze some space

In summary, an extensional model consists of:
\begin{enumerate}
  \item A locally thin reflexive graph internal to CBPV models.
  \item Composition of value and computation relations that form a category with the reflexive relations as identity. Call these categories $\mathcal V_r,\mathcal E_r$
  \item An identity-on-objects functor $\upf : \mathcal V_r \to \mathcal V_f$ taking each value relation to a morphism that universally left-represents it.
  \item An identity-on-objects functor $\dnf : \mathcal E_r^{op} \to \mathcal E_f$ taking each computation relation to a morphism that universally right-represents it.
  \item The CBPV connectives $U,F,\times,\to$ are all \emph{covariant} functorial on relations up to equivalence\footnote{the reflexive graph structure already requires that these functors preserve identity relations}
    \begin{itemize}
    \item $U(dd') \equidyn U(d)U(d')$
    \item $F(cc') \equidyn F(c)F(c')$
    \item $(cc') \to (dd') \equidyn (c \to d)(c' \to d')$
    \item $(c_1c_1') \times (c_2c_2') \equidyn (c_1 \times c_2)(c_1'\times c_2')$
    \end{itemize}
    where $c \equidyn c'$ means $\id \ltsq{c}{c'}\id$ and $\id \ltsq{c'}{c} \id$.
  \item A natural transformation $\mho : 1 \Rightarrow U$ such that
    $\mho \circ ! \ltsq{r(A)}{r(UB)} f$ for any $f : A \to UB$
  \item Distinguished value type $\nat$ with morphisms $z : \mathcal
    V(1,\nat)$ and $s : \mathcal V(\nat,\nat)$.
  \item Distinguished value types $D$ with distinguished relations
    $\iarr{}: U(D \to F D) \rel D$ and $\inat : \nat \rel D$ and $\itimes : D \times D \rel D$
    each satisfying the retraction property $\dnc {\injarr{}}F(\upc{\injarr{}}) \equidyn \id$.
\end{enumerate}

This mathematical structure is sufficient to interpret the theory of
gradual typing in Section~\ref{sec:GTLC}.
\begin{definition}
  Given any extensional gradual typing model, we interpret
  \begin{enumerate}
  \item Each type $A$ as a value type, interpreting the base types and
    $\times$ as their semantic analogues and $\sem {A \ra A'}$ as $U(\sem{A} \to
    F(\sem{A'}))$.
  \item Every type precision derivation $c : A \ltdyn A'$ is
    interpreted as a relation morphism $\sem{c} : \sem{A} \rel \sem{A'}$ in the obvious
    way. Every equivalence axiom $c \equiv c'$ implies that $\sem{c} \equidyn \sem{c'}$.
  \item Every term $\Gamma \vdash M : A$ is interpreted as a morphism
    $\sem{M} : \mathcal V_f (\times\sem{\Gamma},UF\sem{A})$. Upcasts are interpreted as
    $UF(u_{\sem{c}})$ and downcasts as $Ud_{F\sem{c}}$.
  \item If $\Delta \vdash M \ltdyn M' : c$ then $\sem{M}
    \ltsq{\times\sem{\Delta}}{\sem{c}} \sem{M'}$ holds.
  \end{enumerate}
\end{definition}

>>>>

In summary, an extensional model consists of:
\begin{enumerate}
  \item A locally thin reflexive graph internal to CBPV models.
  \item Composition of value and computation relations that form a category with the reflexive relations as identity. Call these categories $\mathcal V_r,\mathcal E_r$
  \item Identity-on-objects functors $\upf : \mathcal V_r \to \mathcal V_f$ and $\dnf : \mathcal E_r^{op} \to \mathcal E_f$ such that every $\upf c$ universally left-represents $c$ and every $\dnf d$ universally represents $d$.
  \item The CBPV connectives $U,F,\times,\to$ are all \emph{covariant} functorial on relations up to equivalence: $U(dd') \equidyn U(d)U(d')$ etc.\footnote{the reflexive graph structure already requires that these functors preserve identity relations}
    %% \begin{itemize}
    %% \item $U(dd') \equidyn U(d)U(d')$
    %% \item $F(cc') \equidyn F(c)F(c')$
    %% \item $(cc') \to (dd') \equidyn (c \to d)(c' \to d')$
    %% \item $(c_1c_1') \times (c_2c_2') \equidyn (c_1 \times c_2)(c_1'\times c_2')$
    %% \end{itemize}
    where $c \equidyn c'$ means $\id \ltsq{c}{c'}\id$ and $\id \ltsq{c'}{c} \id$.
  \item A natural transformation $\mho : 1 \Rightarrow U$ such that
    $\mho \circ ! \ltsq{r(A)}{r(UB)} f$ for any $f : A \to UB$
  \item Distinguished value type $\nat$ with morphisms $z : \mathcal
    V(1,\nat)$ and $s : \mathcal V(\nat,\nat)$.
  \item Distinguished value types $D$ with distinguished relations
    $\iarr{}: U(D \to F D) \rel D$ and $\inat : \nat \rel D$ and $\itimes : D \times D \rel D$
    each satisfying the retraction property $\dnc {\injarr{}}F(\upc{\injarr{}}) \equidyn \id$.
\end{enumerate}

This mathematical structure is sufficient to interpret the theory of
gradual typing in Section~\ref{sec:GTLC}.
\begin{definition}
  Given any extensional gradual typing model, we interpret
  \begin{enumerate}
  \item Each type $A$ as a value type, interpreting the base types and
    $\times$ as their semantic analogues and $\sem {A \ra A'}$ as $U(\sem{A} \to
    F(\sem{A'}))$.
  \item Type precision derivations $c : A \ltdyn A'$ are
    interpreted as relation morphisms $\sem{c} : \sem{A} \rel \sem{A'}$ in the obvious
    way. Every equivalence axiom $c \equiv c'$ implies that $\sem{c} \equidyn \sem{c'}$.
  \item Every term $\Gamma \vdash M : A$ is interpreted as a morphism
    $\sem{M} : \mathcal V_f (\times\sem{\Gamma},UF\sem{A})$. Upcasts are interpreted as
    $UF(u_{\sem{c}})$ and downcasts as $Ud_{F\sem{c}}$.
  \item If $\Delta \vdash M \ltdyn M' : c$ then $\sem{M}
    \ltsq{\times\sem{\Delta}}{\sem{c}} \sem{M'}$ holds.
  \end{enumerate}
\end{definition}
